---
title: "Lab 3: Assignment Plots"
output: 
  github_document: default
  html_notebook: default
always_allow_html: true
---

1. What is the overall change year-on-year? (Hint: plot the temperature distribution for each year as a boxplot) 
2. What is the mean seasonal temperature cycle within a year? 
3. What is the seasonal change at Station 1? Show each year in a different color.
4. What is the seasonal change in temperature at all stations during 2016?


# Initializing

```{r}
setwd("C:/Users/cmitchell/Documents/SeaChangeSemester/2021/Labs/sea-change-semester-data-labs/Lab03/")

library(dplyr)
library(ggplot2)
library(lubridate)
```

# Importing data sets

```{r}
ctdData <- read.csv('DaRTS_CTD_data.csv')
discreteData <- read.csv('DaRTS_discrete_data.csv')
```

# Data wrangling

## Dates

```{r}
ctdData <- mutate(ctdData, Date = ymd(ctdData$Date))

discreteData <- mutate(discreteData, Date = mdy(discreteData$Date))

ctdData <- ctdData %>% 
  mutate(year = year(ctdData$Date),
         month = month(ctdData$Date),
         day = day(ctdData$Date),
         doy = yday(ctdData$Date))

discreteData <- discreteData %>% 
  mutate(year = year(discreteData$Date),
         month = month(discreteData$Date),
         day = day(discreteData$Date),
         doy = yday(discreteData$Date))
```

## Depths

```{r}
ctdData <- ctdData %>% mutate(Depth_m = round(ctdData$Depth_m))

ctdDataBinned <- ctdData %>% 
  group_by(Date, Station, Depth_m) %>% 
  summarize_all(mean, na.rm = TRUE)
```

## Combining datasets

```{r}
combinedData <- ctdDataBinned %>%
  full_join(discreteData)
```

# Plotting

Surface data needed for plotting

```{r}
surfaceData <- combinedData %>%
  filter(Depth_m == 2)
```

## Plot 1

```{r}
ggplot(surfaceData, 
       aes(x=factor(year), y= Temperature_C)) +
  geom_boxplot()
```

## Plot 2

```{r, message=FALSE, results = "hide"}
surfaceData %>% group_by(doy) %>%
  summarize_all(mean) %>%
  ggplot(aes(x=doy, y= Temperature_C)) +
  geom_point() +
  xlim(245,325)
```

## Plot 3

```{r}
surfaceData %>% filter(Station == 1) %>%
  ggplot(aes(x=doy, y= Temperature_C,
             color = factor(year))) +
  geom_point() +
  geom_line() +
  xlim(245,325)
```

## Plot 4

```{r}
surfaceData %>% filter(year == 2016) %>%
  ggplot(aes(x=doy, y=Temperature_C,
             color = factor(Station))) +
  geom_point() +
  geom_line()
```


```{r}
surfaceData %>% filter(year == 2016) %>%
  ggplot(aes(x=doy, y= Station,
             z = Temperature_C)) +
  geom_contour_filled() +
  geom_point() +
  labs(title = 'temperature (C)')
```